Phylotyper: {\it In silico} predictor of molecular subtypes from gene sequences \\ \large Supplementary Information

### 2017-05-01 manuscript/response_to_reviewers.txt

I have two minor comments that are more "food for thought" than required revisions. I am good with the note as is, everything is clear, well cited and concise. For more information it is possible to visit the GitHub page.

"Food for thought" comments:

Although the GitHub page is well laid out and easy enough to follow, please consider including some sort of example to follow along with. As a newcomer to the application it took me a little while to figure out how to get everything exactly right for an initial run. I think a very small example to follow and run that is located early in the GitHub instructions would really help to just get things going.

*Response:
We agree with the reviewer's suggestion that a working example provided for new users will enhance the adoption and usability of Phylotyper.  We have added an "Example" section to the README.md file (which is also used as the landing page for Phylotyper's GitHub repository: https://github.com/superphy/insilico-subtyping/) along with example data that demonstrates how to 1) predict subtypes for schemes already packaged in Phylotyper and 2) how to add and run user-defined subtype schemes.

@@

There lies a slight issue in the "user generated sub-type schemes" in that the results that you get with Phylotyper are highly dependant on the quality of the sub-type scheme provided. Since this tool will likely be used in outbreak situations and analyses, do you have any plans to address the quality of the sub-type scheme that was used? Do you plan on having a section in the GitHub page that allows users to send new sub-type schemes for validation and posting?

*Response*:
The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype predictor. To mitigate poor prediction from unsuitable or inaccurate inputs, we have included in the build pipeline for the local user-submitted subtype feature a series of checks to automatically test the performance of any submitted subtype scheme. These tests 1) verify that the phylogeny of the provided sequences places the same subtypes into phylogenetic groups and 2) that the subtype can be accurately predicted with the given input sequences by running a leave-one-out cross-valiation. The phylogenetic test empirically determines the distribution of inter-patristic distance between the sequences in the tree with the same subtypes. It uses the distribution to identify subtree groupings that place different subtypes in the same clade that are expected to be the same. Users are notified if any unexpected subtype assignments are detected in the input. In our experience, this test was effective at identifying a limited number of anomylous subtype assignments, for example, incorrectly annotated sequences in the larger subtype scheme. In the cross-validation test, each input reference sequence is withheld from the training set and used to test the predictor.  Users are notified if the subtype scheme has a F1-score below 0.9 from the cross-validation test. We found that this test identified subtypes schemes that have poor predictive power due to the subtype being weakly phylogenetically correlated with the input sequences or subtypes being underrepresented in the training set.
While these checks identify major breakdowns in the requirements for an accurate subtype predictor, we note there is no check to identify missing data, specifically, missing subtypes or key subtypes gene sequences that define the range of the subtype sequence space. For any current general subtype predictor (i.e. not predictors that replicate PCR in silico, for instance), performance is dependent on the reference input set. In dealing with missing data, we demonstrate that Phylotyper is significantly more robust than a direct sequence similarity-based predictor. It is less likely to generate a type-I error by when encountering a subtype not found in the training set. The Phylotyper method estimates the uncertainty associated with a subtype prediction. When test sequences that are phylogenetically distant from known reference sequences or in a region of the phylogenetic tree that is rapidly evolving, the Phylotyper score will reflect the degree of uncertaintity. Most often the sequence for a novel or missing subtype will be relatively phylogenetically distant from the other subtype sequences in the tree and its Phylotyper uncertaintity score will reflect this. Only in cases where the novel/missing subbtype is closely phylogenetically related to an existing subtype in the training set, will Phylotyper generate a false positive prediction. This property of Phylotyper makes it more robust when working with incomplete training data.

To enhance the users understanding of the checks that are provided in the build feature and the potential blindspots in creating their own subtype predictors with Phylotyper, we have greatly expanded the methods section and the README provided with the software. The added content describes the two checks, how they work and what they detect. In the README we also offer some guidelines about the inputs needed for an effective subtype predictor. We also included in the manuscript a new analysis that demonstrates how Phylotyper versus a BLAST-based predictor performs when an entire subtype class is removed from the trainging set and used as test input. This analysis and results have been added to the Results and Discussions sections.

The reviewer suggests having a means for users to submit subtype for evaluation and posting to a shared repository. We agree that this would be a helpful system. It will allow microbiologist specialists to submit and improve subtype schemes and increase the 
We have added a section to the GitHub page, subtype scheme will be checked by us and added to the repository or as a separate download.

@@

Sup. Figure 1.
8.      Please plot the legend so that it does not overall the labels
9.      Labels are too close and therefore difficult to read if there were several query allele as an input.

Supplementary table S2:
1.      Four or five E. coli schemes?

Reviewer: 3

Comments to the Author
It's good to see a typing tool that is based on phylogenetic analysis. As public health practitioners still need to compare WGS data to old typing results, this tool will be useful to them.  Moreover, by putting typing into the framework of phylogenetic analysis, this tool can provide an interesting learning/visualization tool for epidemiologists and other public health practitioners to bridge the gap between subtypes (categorical data) and a tree (continuous data). The tool appears to be well implemented.
Some Specific Comments that the authors should address:
Major:
1) Please comment on what happens when an incomplete loci is found in the WGS data. Does the software warn the users of an partial alignment due to incomplete gene?

*Response:
Complete or nearly complete loci sequences are needed for correctly placing the loci in the phylogenetic tree. To prevent incomplete loci being used, we set coverage cutoffs of 90% in the BLAST detection step. No alleles with less than this coverage to one of the reference set sequences will pass the detection step. Users are notified when no loci I have added a description of the requirement for complete loci sequences and the checks we have built into Phylotyper to the Discussion.

2) Please clarify how novel subtypes/alleles are identified and reported.
3) As there is a trend moving to core-genome or whole-genome MLST typing schemes, please comment if Phylotyper could be scaled up to handle that. How would horizontal gene transfer or recombination affect the Phylotyper results?

*Response:
The Phylotyper approach is not likely not applicable to core-genome or whole-genome MLST schemes. The value that Phylotyper provides is that it can take as input gene sequences and return predictions for subtypes based on phenotype or subtypes that cannot be directly inferred from the gene sequences alone. For example, PCR or restriction digestion-based subtyping schemes are an abstraction of the underlying gene sequences and in most cases, cannot be directly inferred from an in silico genome sequence. We have demonstrated the ability of Phylotyper to accurately predict a PCR-based subtype (Stx1 and Stx2) as well as a phenotype-based scheme (O-type:H-type serotype).

Subtypes that are directly identified from the genome sequence do not require Phylotyper's additional prediction step. In MLST, the combination each unique allele sequence defines the type.  Potentially, and this is perhaps what the reviewer is refering to, a specific region of the genome that is smaller than the sequence regions used in an MLST, exists that can be used as a proxy to predict MLST subtype. The benefit in this case, is that the entire genome would not need to be sequenced (or multiple alleles interogated) in order to predict MLST type. There is precedent for this type of approach.  The spa and coa genes in Staphylococcus aureus are used as gene markers to infer phylogenetic and population variation.  The spa and coa types have noted concordance with MLST typing. While this is in interesting potential new application for Phylotyper, it is outside the scope of what we originally designed Phylotyper for.

### 2017-05-01 manuscript/phylotyper_revised.tex
Updated response to reviewers

Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotyping  (Supplementary Table~1).
However, the Phylotyper software aslo has the capability to add new subtype schemes. 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
Checks are built-in to the new subtype pipeline to ensure the assumptions of the Phylotyper approach are not violated.

@@

Performance testing showed that Phylotyper is more robust than an approach based solely on sequence similarity.
Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
By estimating the phylogenetic distribution of each subtype, Phylotyper can more easily identify a novel subtype or accurately classify a novel sequence allele.
The phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting novel alleles.
In comparison, the performance of a sequence similarity approach is highly dependent on the reference database.
This type of approach, there is no inherient mechanism to identify novel alleles.

>>>>

Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotyping  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes. 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
Performance of the Phylotyper method is dependent on the quality of the subtype scheme.
Region phylogeny is predictive of the subtype
Linkage disequalibrium
homologous genes
Correct alignment and construction of phylogeny
Checks are built-in to the new subtype pipeline to ensure the assumptions of the Phylotyper approach are not violated.

@@

Performance testing showed that Phylotyper is more robust than an approach based solely on sequence similarity.
Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make type 1 error when encountering a novel subtype. Our empircal testing confirmed this behavior. 
The rate of false positive classifications when we withheld an entire subtype from the training set and used as it the test set was significantly lower than with a straight-forward

 as one of the subtypes in the database. To validate

or accurately classify a novel sequence allele.
The phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting results.
In comparison, the performance of a sequence similarity approach is highly dependent on the reference database.
This type of approach, there is no inherient mechanism to identify novel alleles.

### 2017-05-01 manuscript/phylotyper_suppl_info_revised.tex
Forgot the new files
Updated response to reviewers

\documentclass[10pt,letterpaper]{article}
%\usepackage[T1]{fontenc}
%\usepackage[utf8]{inputenc}
\usepackage{ragged2e,booktabs,multirow}
\usepackage{caption,capt-of}
\usepackage{graphicx}
\usepackage{array}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

@@

\begin{minipage}{\linewidth}
\captionof{table}{Subtype Schemes in Phylotyper}
\medskip
\begin{tabular}{@{}llll@{}}\toprule Name &
Description & Species & Loci\\\midrule
stx1 & Shiga-toxin 1 subtype & {\it E. coli} & 2 \\
stx2 & Shiga-toxin 1 subtype & {\it E. coli} & 2\\
eae & Intimin subtype & {\it E. coli} & 1\\
flic & H-serotype based on flagellin gene; fliC & {\it E. coli} & 1\\
wz & O-serotype based on the wzy and wzx genes & {\it E. coli} & 2\\\bottomrule
\end{tabular}\par
\bigskip

@@

\bigskip

@@

\begin{minipage}{\linewidth}
\setlength{\tabcolsep}{4pt}
\captionof{table}{Leave-One-Out Cross Validation Results}
\medskip
\begin{tabular}{@{\extracolsep{4pt}}llll>{\centering}m{1.6cm}lll@{}}
\toprule 
\multirow{2}{*}{Scheme} & \multicolumn{4}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
\cline{2-5}\cline{6-8}
& Precision & Recall & F$_{1}$ Score & Run-time (s) $^{1}$ & Precision & Recall & F$_{1}$ Score \\
\midrule
{\it E. coli} Stx1 & 1.00 & 0.94 & 0.97 & 6 & 0.94 & 0.94 & 0.94\\
{\it E. coli} Stx2 & 1.00 & 0.99 & 0.99 & 32 & 0.93 & 0.93 & 0.93\\
{\it E. coli} Intimin & 1.00 & 0.98 & 0.99 & 17 & 0.99 & 0.98 & 0.99\\
{\it E. coli} H-serotype & 0.99 & 0.98 & 0.98 & 16 & 0.96 & 0.85 & 0.90\\
{\it E. coli} O-serotype & 1.00 & 0.61 & 0.75 & 67 & 1.00 & 0.36 & 0.53\\\bottomrule
\end{tabular}\par
\bigskip
Formula:
\begin{enumerate}
\item $Precision = TP / (TP + FP)$
\item $Recall = TP / (TP + FN)$
\item $F_{1}~score = 2*Precision*Recall/(Precision + Recall)$
\end{enumerate}
$TP = True~Positive,~~FP = False~Positive,~~FN = False~Negative$
\par
\bigskip
$^{1}$ Run-time is for an {\it Escherichia coli} genome containing a non-identical subtype gene which triggers the full phylotyper pipeline.

>>>>

\documentclass[11pt,letterpaper]{article}
%\usepackage[T1]{fontenc}
%\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{ragged2e,booktabs,multirow}
\usepackage{caption,capt-of}
\usepackage{graphicx}
\usepackage{array}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

@@

~

\begin{minipage}{\linewidth}
\centering
\captionof{table}{Subtype Schemes in Phylotyper}
\medskip
\begin{tabular}{@{}llll@{}}\toprule Name &
Description & Species & Loci\\\midrule
stx1 & Shiga-toxin 1 subtype & {\it E. coli} & 2 \\
stx2 & Shiga-toxin 1 subtype & {\it E. coli} & 2\\
eae & Intimin subtype & {\it E. coli} & 1\\
flic & H-serotype based on flagellin gene; fliC & {\it E. coli} & 1\\
wz & O-serotype based on the wzy and wzx genes & {\it E. coli} & 2\\\bottomrule
\end{tabular}\par

@@

@@

%\captionsetup{justification=RaggedRight}

\begin{minipage}{\linewidth}
\centering
\setlength{\tabcolsep}{4pt}
\captionof{table}[justification=RaggedRight]{Leave-One-Out Cross Validation Results}
\medskip
\begin{tabular}{@{\extracolsep{4pt}}llll>{\centering}m{1.6cm}lll@{}}
\toprule 
\multirow{2}{*}{Scheme} & \multicolumn{4}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
\cline{2-5}\cline{6-8}
& Precision & Recall & F$_{1}$ Score & Run-time (s) $^{1}$ & Precision & Recall & F$_{1}$ Score \\
\midrule
{\it E. coli} Stx1 & 1.00 & 0.94 & 0.97 & 6 & 0.94 & 0.94 & 0.94\\
{\it E. coli} Stx2 & 1.00 & 0.99 & 0.99 & 32 & 0.93 & 0.93 & 0.93\\
{\it E. coli} Intimin & 1.00 & 0.98 & 0.99 & 17 & 0.99 & 0.98 & 0.99\\
{\it E. coli} H-serotype & 0.99 & 0.98 & 0.98 & 16 & 0.96 & 0.85 & 0.90\\
{\it E. coli} O-serotype & 1.00 & 0.61 & 0.75 & 67 & 1.00 & 0.36 & 0.53\\\bottomrule
\end{tabular}\par
\bigskip
\raggedright
Formula:
\begin{enumerate}
\item $Precision = TP / (TP + FP)$
\item $Recall = TP / (TP + FN)$
\item $F_{1}~score = 2*Precision*Recall/(Precision + Recall)$
\end{enumerate}
$TP = True~Positive,~~FP = False~Positive,~~FN = False~Negative$
\par
\bigskip
$^{1}$ Run-time is for an {\it Escherichia coli} genome containing a unique gene which triggers the full phylotyper pipeline.

\begin{minipage}{\linewidth}
\centering
\captionof{table}{V-typer Performance}
\medskip
\begin{tabular}{@{}lll@{}}\toprule
NCBI Accession & Stx Subtype & V-typer Prediciton$^{1}$ \\\midrule
CP015020.1 & 2d & 2d & N \\
stx2 & Shiga-toxin 1 subtype & {\it E. coli} & 2\\
eae & Intimin subtype & {\it E. coli} & 1\\
flic & H-serotype based on flagellin gene; fliC & {\it E. coli} & 1\\
wz & O-serotype based on the wzy and wzx genes & {\it E. coli} & 2\\\bottomrule
\end{tabular}\par
\bigskip
\raggedright
$^{1}$ N=Negative (no results returned)
